---
title: "Cortical Surface Data Format Conversion"
output: html_notebook
---

# Code for converting BISCUIT cortical surface outputs to Freesurfer format for data analysis in R
Under development Jan. 2025 by Meaghan Perdue

```{r}
library(dplyr)
library(freesurferformats)
library(knitr)

setwd("/Users/meaghan/Projects/cortical-analysis/src")

```
## Convert OASIS45 template surfaces to Freesurfer format
This step needs to be run only once, so you can just pull the files from Rundle
Oasis45 template .vtk and converted .mgh files can be found on Rundle at: /Volumes/catherine_team/Project_Folders/MaCRUISE/oasis45
```{r}
#read OASIS45 left hemisphere inflated input
oasis.lh.inflated<-read.fs.surface.vtk('/Users/meaghan/Projects/cortical-analysis/oasis45/lh.inflated.vtk')

#convert to freesurfer .mgh format
write.fs.surface('/Users/meaghan/Projects/cortical-analysis/oasis45/lh.inflated.mgh', oasis.lh.inflated$vertices, oasis.lh.inflated$faces, format = 'bin')

#read OASIS45 right hemisphere inflated input
oasis.rh.inflated<-read.fs.surface.vtk('/Users/meaghan/Projects/cortical-analysis/oasis45/rh.inflated.vtk')

#convert to freesurfer .mgh format
write.fs.surface('/Users/meaghan/Projects/cortical-analysis/oasis45/rh.inflated.mgh', oasis.rh.inflated$vertices, oasis.rh.inflated$faces, format = 'bin')
```

## Read BISCUIT template-registered scalar data and write to freesurfer format
set paths to macruise and subdir according to your system set-up and project folder
"sublist.txt" is a 2 column tab-separated list with Col1=MaCRUISE folder name (sub_Session) and Col2=desired Output folder name (can be the same or not) \
"sublist.txt" should be in the same directory as this script, otherwise include full path in the read.table function \
outputs are freesurfer-format scalar files in fsaverage space

```{r}

list<-read.table("sublist.txt")

for (i in 1:length(list$V1)) {
    #set macruise folder ID
    macid<-list$V1[i]
    
    #set output folder ID
    out<-list$V2[i]
    
    # set path to subject's macruise output
    macruise<-Sys.glob(file.path("/Volumes/catherine_team/Project_Folders/MaCRUISE/PS_MaCRUISE1mm/1_Processed_Output", macid, "*biscuit*", "stat"))
    
    # create subject output directory for freesurfer formatted data
    subdir<-file.path("/Users/meaghan/Projects/cortical-analysis/data", out)
    dir.create(subdir)
    
    # Cortical Thickness
    lh.reg.ct.vals<-read.table(file.path(macruise, 'lh.reg.thickness.txt'))$V1
    
    write.fs.curv(file.path(subdir,'lh.reg.ct'), lh.reg.ct.vals)
    
    rh.reg.ct.vals<-read.table(file.path(macruise,'rh.reg.thickness.txt'))$V1
    
    write.fs.curv(file.path(subdir, 'rh.reg.ct'), rh.reg.ct.vals)
    
    # Surface Area
    lh.reg.sa.vals<-read.table(file.path(macruise,'lh.mid.reg.area.txt'))$V1
    
    write.fs.curv(file.path(subdir, 'lh.reg.sa'), lh.reg.sa.vals)
    
    rh.reg.sa.vals<-read.table(file.path(macruise,'rh.mid.reg.area.txt'))$V1
    
    write.fs.curv(file.path(subdir, 'rh.reg.sa'), rh.reg.sa.vals)
    
    # Local Gyrification Index
    lh.reg.lgi.vals<-read.table(file.path(macruise,'lh.gray.reg.LGI.txt'))$V1
    
    write.fs.curv(file.path(subdir, 'lh.reg.lgi'), lh.reg.lgi.vals)
    
    rh.reg.lgi.vals<-read.table(file.path(macruise,'rh.gray.reg.LGI.txt'))$V1
    
    write.fs.curv(file.path(subdir, 'rh.reg.lgi'), rh.reg.lgi.vals)
    
    # Sulcal Depth
    lh.reg.sd.vals<-read.table(file.path(macruise,'lh.gray.reg.SD.txt'))$V1
    
    write.fs.curv(file.path(subdir,'lh.reg.sd'), lh.reg.sd.vals)
    
    rh.reg.sd.vals<-read.table(file.path(macruise,'rh.gray.reg.SD.txt'))$V1
    
    write.fs.curv(file.path(subdir, 'rh.reg.sd'), rh.reg.sd.vals)
    
    # Shape Complexity Index
    lh.reg.sci.vals<-read.table(file.path(macruise,'lh.mid.reg.SCI.txt'))$V1
    
    write.fs.curv(file.path(subdir, 'lh.reg.sci'), lh.reg.sci.vals)
    
    rh.reg.sci.vals<-read.table(file.path(macruise,'rh.mid.reg.SCI.txt'))$V1
    
    write.fs.curv(file.path(subdir, 'rh.reg.sci'), rh.reg.sci.vals)

}

```


## Resample data from oasis45 space to fsaverage6 space
VertexWiseR requires data to be resampled to fsaverage5 or fsaverage6 (40692 vertices per hemisphere) space \
Runs resampling in freesurfer using mris_apply_reg tool (freesurfer must be installed on local machine, make sure FREESURFER_HOME and SUBJECTS_DIR paths are correct) \
copy or link the oasis45 template folder to your WORK_DIR \
subs is list of output folder names from the previous step, must be in quotes \
Outputs (.mgz files) can be viewed in freeview as an overlay on the fsaverage6 surface
```{bash}
export FREESURFER_HOME=/Applications/freesurfer/7.4.1
export SUBJECTS_DIR=/Applications/freesurfer/7.4.1/subjects
#set working directory for outputs
export WORK_DIR=/Users/meaghan/Projects/cortical-analysis

subs=(
"10117_Session25"
"10117_Session26"
"10117_Session27"
"10117_Session28"
"10136_Session12"
"10136_Session16"
"10136_Session17"
"10136_Session18"
"10136_Session19"
"10136_Session20"
"10136_Session21"
"10136_Session22"
"10166_Session5"
"10166_Session6"
"10166_Session7"
"10166_Session8"
"10166_Session9"
"10166_Session10"
"10166_Session11"
"10162_Session1"
)
    
metricsleft=(
    "lh.reg.ct"
    "lh.reg.lgi"
    "lh.reg.sa"
    "lh.reg.sd"
    "lh.reg.sci"
    )
    
metricsright=(
    "rh.reg.ct"
    "rh.reg.lgi"
    "rh.reg.sa"
    "rh.reg.sd"
    "rh.reg.sci"
    )


for i in ${subs[@]}; do
  for j in ${metricsleft[@]}; do
    # Left hemisphere
    mris_apply_reg --i $WORK_DIR/data/${i}/${j} \
      --o  $WORK_DIR/data/${i}/${j}.fs6.mgz \
      --streg $WORK_DIR/oasis45/lh.inflated.mgh $SUBJECTS_DIR/fsaverage6/surf/lh.inflated 
    done
  for k in ${metricsright[@]}; do
    # Right hemisphere
    mris_apply_reg --i $WORK_DIR/data/${i}/${k} \
      --o  $WORK_DIR/data/${i}/${k}.fs6.mgz \
      --streg $WORK_DIR/oasis45/rh.inflated.mgh $SUBJECTS_DIR/fsaverage6/surf/rh.inflated 
    done
  done


```

## Prepare data for Vertex Wise Analysis with VertexWiseR
Concatenate left-hemisphere and right-hemisphere data per subject/session \
Create data matrix for all subjects/sessions to be included in analysis 
*This section under development as of 24 Jan 2025, currently set up to test a few subjects, need to set up batch processing*
```{r}
library(freesurferformats)
library(VertexWiseR)

# read resampled thickness data
lh.thickness.fs6 <- read.fs.mgh("/Users/meaghan/Projects/cortical-analysis/data/10046_Session7/lh.thickness.fsa6.mgz")
rh.thickness.fs6 <- read.fs.mgh("/Users/meaghan/Projects/cortical-analysis/data/10046_Session7/rh.thickness.fsa6.mgz")

# per-participant concatenated scalar data lh, rh
s1.bilat.thickness.fs6<-c(lh.thickness.fs6, rh.thickness.fs6)
s2.bilat.thickness.fs6<-c(lh.thickness.fs6, rh.thickness.fs6)
s3.bilat.thickness.fs6<-c(lh.thickness.fs6, rh.thickness.fs6)

# extract surface data to VertexWiseR matrix format: 1 row per subject containing vertex-wise measurements lh followed by rh
matrix<-matrix(c(s1.bilat.thickness.fs6, s2.bilat.thickness.fs6, s3.bilat.thickness.fs6), nrow=3, byrow=TRUE)


# save matrix to rds
saveRDS(matrix, "/Users/meaghan/Projects/cortical-analysis/data/test/thickness_matrix_fs6.rds")

# smooth suface data
# smoothing kernel should be adjusted depending on the appropriate FWHM for each metric
thickness_smoothed<-smooth_surf(matrix, 10, VWR_check = FALSE)



```

# Run vertex-wise analysis
Define covariates and models for vertex-wise analysis \
VertexWiseR can accommodate linear models and mixed-effects models \
Random Field Theory (RFT) cluster correction or Threshold-Free Cluster Enhancement (TFCE) options available \
VertexWiseR documentation: <https://github.com/CogBrainHealthLab/VertexWiseR>
*This section under development as of 24 Jan 2025, currently set up to test package functioning with random numbers, models should be set up according to VertexWiseR documentation*
```{r}
# define covariates  
all_pred=data.frame(age = c(7, 8, 9))
head(all_pred)

# run the model
model.thickness=TFCE_vertex_analysis(model = all_pred$age, contrast=all_pred$age, surf_data = thickness_smoothed)

# write T-stat files to surface scalar files for viewing in freeview (left and right hemispheres written separately by index)
write.fs.curv('/Users/meaghan/Projects/cortical-analysis/data/test/tfce.tstat.lh', model.thickness$t_stat[1:40962])
write.fs.curv('/Users/meaghan/Projects/cortical-analysis/data/test/tfce.tstat.rh', model.thickness$t_stat[40963:81924])


# threshold results
results=TFCE_threshold(model.thickness, p=0.05, atlas=1)
print(results$cluster_level_results)

```

## Optional code for individual subject conversion/visualization
### Read BISCUIT gray matter surfaces and write to freesurfer format
We don't really need to do this because we'll be working with the resampled surface data, but useful for visualization of individual subject data in freeview
```{r}
lh.pial<-read.fs.surface.vtk('/Volumes/catherine_team/Project_Folders/MaCRUISE/PS_MaCRUISE1mm/1_Processed_Output/10006_PS20_012_Session7/Lebel_Peds-x-10006-x-10006_PS20_012_Session7-x-biscuit_mc_v2-x-5354db8f-b70a-4597-9fb7-315acf79ff7c/surf/lh.gray.vtk')

write.fs.surface('/Users/meaghan/Projects/cortical-analysis/data/test/Rlh.pial', lh.pial$vertices, lh.pial$faces, format = 'bin')

rh.pial<-read.fs.surface.vtk('/Volumes/catherine_team/Project_Folders/MaCRUISE/PS_MaCRUISE1mm/1_Processed_Output/10006_PS20_012_Session7/Lebel_Peds-x-10006-x-10006_PS20_012_Session7-x-biscuit_mc_v2-x-5354db8f-b70a-4597-9fb7-315acf79ff7c/surf/rh.gray.vtk')

write.fs.surface('/Users/meaghan/Projects/cortical-analysis/data/test/Rrh.pial', rh.pial$vertices, rh.pial$faces, format = 'bin')

```

### Read BISCUIT scalar data and write to freesurfer format
Again, this is just for visualization of each participant's data prior to registration/resampling
```{r}
lh.lgi.vals<-read.table('/Volumes/catherine_team/Project_Folders/MaCRUISE/PS_MaCRUISE1mm/1_Processed_Output/10006_PS20_012_Session7/Lebel_Peds-x-10006-x-10006_PS20_012_Session7-x-biscuit_mc_v2-x-5354db8f-b70a-4597-9fb7-315acf79ff7c/quant/lh.gray.LGI.txt')$V1

write.fs.curv('/Users/meaghan/Projects/cortical-analysis/data/test/lh.lgi', lh.lgi.vals)

rh.lgi.vals<-read.table('/Volumes/catherine_team/Project_Folders/MaCRUISE/PS_MaCRUISE1mm/1_Processed_Output/10006_PS20_012_Session7/Lebel_Peds-x-10006-x-10006_PS20_012_Session7-x-biscuit_mc_v2-x-5354db8f-b70a-4597-9fb7-315acf79ff7c/quant/rh.gray.LGI.txt')$V1

write.fs.curv('/Users/meaghan/Projects/cortical-analysis/data/test/rh.lgi', rh.lgi.vals)

```
