---
title: "Cortical Surface Data Format Conversion"
output: html_notebook
---

# Code for converting BISCUIT cortical surface outputs to Freesurfer format for data analysis in R
Under development Jan. 2025 by Meaghan Perdue

```{r}
library(dplyr)
library(freesurferformats)
library(knitr)

setwd("/Users/meaghan/Projects/cortical-analysis/src")


```

## Read BISCUIT gray matter surfaces and write to freesurfer format
We don't really need to do this because we'll be working with the resampled surface data, but useful for visualization of individual subject data
```{r}
lh.pial<-read.fs.surface.vtk('/Volumes/catherine_team/Project_Folders/MaCRUISE/PS_MaCRUISE1mm/1_Processed_Output/10006_PS20_012_Session7/Lebel_Peds-x-10006-x-10006_PS20_012_Session7-x-biscuit_mc_v2-x-5354db8f-b70a-4597-9fb7-315acf79ff7c/surf/lh.gray.vtk')

write.fs.surface('/Users/meaghan/Projects/cortical-analysis/data/test/Rlh.pial', lh.pial$vertices, lh.pial$faces, format = 'bin')

rh.pial<-read.fs.surface.vtk('/Volumes/catherine_team/Project_Folders/MaCRUISE/PS_MaCRUISE1mm/1_Processed_Output/10006_PS20_012_Session7/Lebel_Peds-x-10006-x-10006_PS20_012_Session7-x-biscuit_mc_v2-x-5354db8f-b70a-4597-9fb7-315acf79ff7c/surf/rh.gray.vtk')

write.fs.surface('/Users/meaghan/Projects/cortical-analysis/data/test/Rrh.pial', rh.pial$vertices, rh.pial$faces, format = 'bin')

```

## Read BISCUIT scalar data and write to freesurfer format
Again, this is just for visualization of each participant's data prior to registration/resampling
```{r}
lh.lgi.vals<-read.table('/Volumes/catherine_team/Project_Folders/MaCRUISE/PS_MaCRUISE1mm/1_Processed_Output/10006_PS20_012_Session7/Lebel_Peds-x-10006-x-10006_PS20_012_Session7-x-biscuit_mc_v2-x-5354db8f-b70a-4597-9fb7-315acf79ff7c/quant/lh.gray.LGI.txt')$V1

write.fs.curv('/Users/meaghan/Projects/cortical-analysis/data/test/lh.lgi', lh.lgi.vals)

rh.lgi.vals<-read.table('/Volumes/catherine_team/Project_Folders/MaCRUISE/PS_MaCRUISE1mm/1_Processed_Output/10006_PS20_012_Session7/Lebel_Peds-x-10006-x-10006_PS20_012_Session7-x-biscuit_mc_v2-x-5354db8f-b70a-4597-9fb7-315acf79ff7c/quant/rh.gray.LGI.txt')$V1

write.fs.curv('/Users/meaghan/Projects/cortical-analysis/data/test/rh.lgi', rh.lgi.vals)

```

## Read BISCUIT template-registered scalar data and write to freesurfer format
outputs are freesurfer-format scalar files in fsaverage space 

```{r}

list<-read.table("sublist.txt")

for (i in 1:length(list$V1)) {


    #set macruise folder ID
    macid<-list$V1[i]
    
    #set output folder ID
    out<-list$V2[i]
    
    # set path to subject's macruise output
    macruise<-Sys.glob(file.path("/Volumes/catherine_team/Project_Folders/MaCRUISE/PS_MaCRUISE1mm/1_Processed_Output", macid, "*biscuit*", "stat"))
    
    # create subject output directory for freesurfer formatted data
    subdir<-file.path("/Users/meaghan/Projects/cortical-analysis/data", out)
    dir.create(subdir)
    
    # Cortical Thickness
    lh.reg.ct.vals<-read.table(file.path(macruise, 'lh.reg.thickness.txt'))$V1
    
    write.fs.curv(file.path(subdir,'lh.reg.ct'), lh.reg.ct.vals)
    
    rh.reg.ct.vals<-read.table(file.path(macruise,'rh.reg.thickness.txt'))$V1
    
    write.fs.curv(file.path(subdir, 'rh.reg.ct'), rh.reg.ct.vals)
    
    # Surface Area
    lh.reg.sa.vals<-read.table(file.path(macruise,'lh.mid.reg.area.txt'))$V1
    
    write.fs.curv(file.path(subdir, 'lh.reg.sa'), lh.reg.sa.vals)
    
    rh.reg.sa.vals<-read.table(file.path(macruise,'rh.mid.reg.area.txt'))$V1
    
    write.fs.curv('/Users/meaghan/Projects/cortical-analysis/data/test/rh.reg.sa', rh.reg.sa.vals)
    
    # Local Gyrification Index
    lh.reg.lgi.vals<-read.table(file.path(macruise,'lh.gray.reg.LGI.txt'))$V1
    
    write.fs.curv(file.path(subdir, 'lh.reg.lgi'), lh.reg.lgi.vals)
    
    rh.reg.lgi.vals<-read.table(file.path(macruise,'rh.gray.reg.LGI.txt'))$V1
    
    write.fs.curv(file.path(subdir, 'rh.reg.lgi'), rh.reg.lgi.vals)
    
    # Sulcal Depth
    lh.reg.sd.vals<-read.table(file.path(macruise,'lh.gray.reg.SD.txt'))$V1
    
    write.fs.curv(file.path(subdir,'lh.reg.sd'), lh.reg.sd.vals)
    
    rh.reg.sd.vals<-read.table(file.path(macruise,'rh.gray.reg.SD.txt'))$V1
    
    write.fs.curv(file.path(subdir, 'rh.reg.sd'), rh.reg.sd.vals)
    
    # Shape Complexity Index
    lh.reg.sci.vals<-read.table(file.path(macruise,'lh.mid.reg.SCI.txt'))$V1
    
    write.fs.curv(file.path(subdir, 'lh.reg.sci'), lh.reg.sci.vals)
    
    rh.reg.sci.vals<-read.table(file.path(macruise,'rh.mid.reg.SCI.txt'))$V1
    
    write.fs.curv(file.path(subdir, 'rh.reg.sci'), rh.reg.sci.vals)

}

```


## Resample data from fsaverage space to fsaverage6 space
VertexWiseR requires data to be resampled to fsaverage5 or fsaverage6 (40692 vertices per hemisphere) space
Runs resampling in freesurfer using mris_apply_reg tool (freesurfer must be installed on local machine, make sure FREESURFER_HOME and SUBJECTS_DIR paths are correct)
```{bash}
export FREESURFER_HOME=/Applications/freesurfer/7.4.1
export SUBJECTS_DIR=/Applications/freesurfer/7.4.1/subjects

# Left hemisphere
mris_apply_reg --i /Users/meaghan/Projects/cortical-analysis/data/10046_Session7/lh.reg.ct \
--o  /Users/meaghan/Projects/cortical-analysis/data/10046_Session7/lh.thickness.fsa6.mgz \
--streg $SUBJECTS_DIR/fsaverage/surf/lh.sphere.reg $SUBJECTS_DIR/fsaverage6/surf/lh.sphere.reg 

# Right hemisphere
mris_apply_reg --i /Users/meaghan/Projects/cortical-analysis/data/10046_Session7/rh.reg.ct \
--o  /Users/meaghan/Projects/cortical-analysis/data/10046_Session7/rh.thickness.fsa6.mgz \
--streg $SUBJECTS_DIR/fsaverage/surf/rh.sphere.reg $SUBJECTS_DIR/fsaverage6/surf/rh.sphere.reg 


```
## Prepare data for Vertex Wise Analysis with VertexWiseR

```{r}
library(freesurferformats)
library(VertexWiseR)

# read resampled thickness data
lh.thickness.fs6 <- read.fs.mgh("/Users/meaghan/Projects/cortical-analysis/data/10046_Session7/lh.thickness.fsa6.mgz")
rh.thickness.fs6 <- read.fs.mgh("/Users/meaghan/Projects/cortical-analysis/data/10046_Session7/rh.thickness.fsa6.mgz")

s1.bilat.thickness.fs6<-c(lh.thickness.fs6, rh.thickness.fs6)
s2.bilat.thickness.fs6<-c(lh.thickness.fs6, rh.thickness.fs6)
s3.bilat.thickness.fs6<-c(lh.thickness.fs6, rh.thickness.fs6)

matrix<-matrix(c(s1.bilat.thickness.fs6, s2.bilat.thickness.fs6, s3.bilat.thickness.fs6), nrow=3, byrow=TRUE)

input<-"/Users/meaghan/Projects/cortical-analysis/data/test"

# extract surface data to VertexWiseR matrix format: 1 row per subject containing vertex-wise measurements lh followed by rh
#sample<-readRDS(file="~/Downloads/SPRENG_CTv_site1.rds")

#set macruise folder ID
   #cocatenate subjects' hemispheres data 
    thickness<-c(lh.reg.ct.vals, rh.reg.ct.vals)
    thickness2<-c(lh.reg.ct.vals, rh.reg.ct.vals)
    thickness3<-c(lh.reg.ct.vals, rh.reg.ct.vals)
    
#pull multiple subjects into matrix
    matrix<-matrix(c(thickness, thickness2, thickness3), nrow=3, byrow=TRUE)

# save matrix to rds
saveRDS(matrix, "/Users/meaghan/Projects/cortical-analysis/data/test/thickness_matrix_fs6.rds")

# smooth suface
thickness_smoothed<-smooth_surf(matrix, 10, VWR_check = FALSE)



```

# Run vertex-wise analysis
```{r}
# define covariates  
all_pred=data.frame(age = c(7, 8, 9))
head(all_pred)

# run the model
model.thickness=TFCE_vertex_analysis(model = all_pred$age, contrast=all_pred$age, surf_data = thickness_smoothed)

# write T-stat files to surface scalar files for viewing in freeview (left and right hemispheres written separately by index)
write.fs.curv('/Users/meaghan/Projects/cortical-analysis/data/test/tfce.tstat.lh', model.thickness$t_stat[1:40962])
write.fs.curv('/Users/meaghan/Projects/cortical-analysis/data/test/tfce.tstat.rh', model.thickness$t_stat[40963:81924])


# threshold results
results=TFCE_threshold(model.thickness, p=0.05, atlas=1)
print(results$cluster_level_results)

```
